The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-29 paper.qmd
EDA section

## Can we have uncertinty visualisations for EDA?
A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another. This problem becomes much more explicit through the use of the VSUP.

@@

The ramifications of this problem is seen repeatedly in the uncertaint visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.

>>>>

## Can we have uncertinty visualisations for EDA?
A slight problem with uncertainty visualisation is that the uncertainty and the purpose of visualisation are somewhat at odds with one another. The VSUP highlights this problem quite well.

There are two primary motivations behind visualisation, communication and exploratory data analysis (EDA). Communication involves identifying a signal we want to communicate and designing a visualisation that best conveys that, while EDA involves creating a versatile visualisation using it to extract several signals. It is not uncommon for authors to express a desire for uncertainty visualisations that perform signal-suppression in visualisations made for EDA [@Sarma2024; @Griethe2006], however it is not clear that these uncertainty visualisations *exist at all*.

The impossible task of making this graphic is easy to see with a simple example. Let's say we have a graphic that depicts a set of coefficients from a linear regression and the value of the coefficient is shown using a single colour. We do a series of t-tests on these estimates, and all of them fail to reject the null hypothesis that they are different from 0. This would indicate that each value needs to be visualised using a single color that is indistinguishable from the colour assigned to 0. However, we also perform an f-test on our model, and this test *is* rejected. This test indicates that the set of estimates should not be visually indistinguishable from 0. This is clearly a contradiction.

This t-test, f-test example highlights a fundamental problem with the VSUP and even the bivariate map. When we blend colors to make them visually indistinguishable, we need to decide at what level of *uncertainty* to blend these colors together. There is no intrinsic mapping of "uncertainty", so we need to make a decision as to how much "suppressing" the visualisation does. We did this implicitly in the bivarate map by deciding how much "variance"

- we could consider this a power thing?

Using the methods discussed thus far, where each cell is colours by a single value, it would be impossible to succeed in this method. However, this may not be true if we change our approach entirely.

@@

The ramifications of this problem is seen repeatedly in the uncertainty visualisation literature. A survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] ignore the problem get around this definition issue by focusing narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values, but ignore these concepts when translating that uncertainty to a visualisation. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, and information theory tries to quantify uncertainty using the idea of entropy, but the disconnect between the broad concept of uncertainty and what we can reliably quantify remains a problem.
